---
title: "Acquiring Static Data in R"
subtitle: "Bytes & Bridges - Data Engineering"
author: "Ricardo Mendes"
date: "2024-04-16"
output:
  html_notebook: default
editor_options: 
  markdown: 
    wrap: sentence
include-before: '`\newpage{}`{=latex}'
---

```{r knitr-options, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff=50),
  warning = FALSE, 
  message = FALSE,  
  fig.align="center", 
  out.width="70%")

options(warn=-1)
```


# **Acquiring Static Data**

In this project, let's explore how to use Python to obtain static data for any purpose you need. All the files used in this project are in the same folder as this file.

## **Excel File**

Excel files are widely used in companies and can be an easy to create file when you need to import structured data.

```{r read-excel}
if(!require(xlsx)) install.packages("xlsx", INSTALL_opts=c("--no-multiarch"))
library("xlsx")

car_list <- read.xlsx("Car List.xlsx", sheetIndex=1)
head(car_list)

```

## **CSV File**

CSV files are also used because they are easy to interoperate and consume. Reading them in Python is also pretty straightforward:

``` {r read-csv}
# Read a CSV file and show the first 5 rows
countries <- read.csv("Countries.csv", sep = ";")
head(countries)

```

## **JSON File**

Another type of file that is very common is the JSON file. Here is how you can read specific values within it:

```{r read-json}
# Read the JSON file
user_config <- jsonlite::fromJSON("UserConfig.json")

# Print the values within the configuration
cat("Username:", user_config$username, "\n")
cat("Language:", user_config$language, "\n")
cat("Fullscreen:", user_config$fullscreen, "\n")
```

## **Image File**

It is often necessary to read image files. They can be used as icons, logos, or for any other purpose. Check below how to do that:

``` {r read-image}

if(!require(jpeg)) install.packages("jpeg")
library("jpeg")

# Read the image file
image_path = "Image.jpeg"
image <- readJPEG(image_path, native = TRUE)

# Open the file in a separate window:
plot(0:1, 0:1, type = "n", ann = FALSE, axes=FALSE)

# Show the image inline
rasterImage(image, 0, 0, 1, 1)

```

## **Web Pages**

In some cases, you might want to read information directly from a web page, instead of selecting it, creating a file, and then reading the file.
The code below shows one of the many ways in which you can accomplish that by reading the list of the busiest airports in 2022:

``` {r read-webpage}
# Load necessary libraries
library(rvest)  # For web scraping
library(dplyr)  # For data manipulation

# URL of the Wikipedia page
url <- "https://en.wikipedia.org/wiki/List_of_busiest_airports_by_passenger_traffic"

# Read the HTML content from the URL
webpage <- read_html(url)

# Extract the table containing the airport data (second table on the page)
airport_table <- html_nodes(webpage, "table.wikitable.sortable")[2]

# Convert the table to a dataframe
df_airports <- airport_table %>%
  html_table(fill = TRUE) %>%
  as.data.frame()

# Rename columns for clarity
colnames(df_airports) <- c("Rank", "Airport", "Location", "Country", "Code", "Total_Passengers", "Rank_Change", "Percent_Change")

# Select the most relevant columns
df_airports <- df_airports %>%
  select(Rank, Airport, Location, Total_Passengers)

# Display the resulting dataframe
df_airports
```



